### 2020-06-12 manuscript/internship_report.Rmd
update report

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

@@

## Eye movement metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user (e.g., the distance between the mean position in a 100 ms window before and after each sample, see Olsson, 2007, and Zemblys et al.). This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics. Belonging to the most basic metrics, which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [similar to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013]. They are specifically distinguished by their change in direction clustered around 180° [@Pekkanen2017]. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].

@@

(ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s). Line color indicates whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue). For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3. Full line segment marks where potential oscillations are filtered out. Dashed line segment illustrates where potentially random change in angle is filtered out.

@@

Instead of filtering before computing the metrics, I decided to combine both in one step. Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration). @Houpt2018 compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm. Similarly, @Holmqvist2010 use a Savitzky-Golay (SG) filter @Savitzky1964 to estimate velocity and acceleration from a gaze signal. Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods. They follow a similar trend, but the SG velocity signal is less noisy and acceleration signal is consequently lower than the Laplace acceleration signal. Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events. To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the backward difference (see Technical Appendix).

(ref:filter-comparison) Example data from Andersson et al. (2017) displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s). Line colors indicate which filter has been used to derive the signal. Savitzky-Golay filters were applied with order three and length five (corresponding to 10 ms).

@@

## The generative model
The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM). It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits. Thus, users can choose wether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth puruits.  
In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model [@Visser2019]. In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal. The response variables are treated as conditionally independent on the states. I acknowledge that conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated). This step was merely taken to keep the HMM simple and identifyable.  
Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them). However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation) and due to noise above zero. Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean. Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties. In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose. It has to be noted that the gamma was choosen because of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution. Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$) which should lead to rather symmetric distributions. Moreover, the uniform distribution can distinguish fixations from the other events.  
The initial state model and the transition model used multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.

## Missing data and blinks
HMMs can estimate parameters and hidden states despite missing data. They either treat data as missing at random or as missing dependent on states [@Visser2019]. In gazeHMM, I assume data to be missing at random. Most of the missing data in eye movement classification are due to blinks. The user can indicate in gazHMM which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, I decided to remove them heuristically. Before classifying the samples, gazeHMM sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.  
When evaluating the likelihood of the HMM given parameters and the joint density of response distributions for missing data, the model integrates over all possible values. The resulting density for missing samples is therefore set to one [@Visser2019].

@@

apa_table(data.frame("Event" = rep(c("Fixation", "Saccade", "PSO", "Smooth pursuit"), each = 3),
                     "Resp variable" = rep(c("Velocity", "Acceleration", "Angle"), 4),
                     "Range" = c("0-50", "0-50", "uniform", "50-1000", "50-500", "~0", "20-100", "10-90", paste("~", expression(pi), sep = ""), "20-100", "0-30", "~0")), 
          caption = "(ref:sim-ranges)",
          note = "(ref:sim-ranges-note)",
          escape = F)

@@

```{r, sim-parameters}

sim.pars.df <- data.frame(State = c(rep("1-4", 3), as.character(rep(1:4, each = 6))),
                          #Parameter = c(""),
                          Interval = c("-", "[.01,.99]", "1-a(i=j)/(k-1)", 
                                       "[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
                                       "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
                                       "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
                                       "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]"),
                          Default = c("1/k", "0.9", "0.1/(k-1)", 
                                      "3", "0.35", "3", "0.25", "0", paste(expression(pi), sep = ""),
                                      "3", "10", "3", "3", "0", "1",
                                      "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
                                      "3", "1", "3", "0.15", "0", "1"),
                          Description = c("Initial state probability for starting in state *i*",
                                          "Transition probability for staying in the previous state *i*",
                                          "Transition probability for switching to from state *i* to a different state *j*",
                                          "Shape parameter of the velocity gamma distribution",
                                          "Scale parameter of the velocity gamma distribution",
                                          "Shape parameter of the acceleration gamma distribution",
                                          "Scale parameter of the acceleration gamma distribution",
                                          "Minimum of the uniform distribution",
                                          "Maximum of the uniform distribution",
                                          "", "", "", "",
                                          "Mean parameter of the von-Mises distribution",
                                          "Concentration parameter of the von Mises distribution",
                                          rep("", 12)))

sim.pars.names <- c(expression(rho[i]), expression(a[i==j]), expression(a[i!=j]),
                   expression(alpha[vel]))

rownames(sim.pars.df)[1:4] <- sim.pars.names

apa_table(sim.pars.df, row.names = T)

@@

```{r sim results and helper functions, include=FALSE}

@@

## Validation
### Model comparison

@@

To test whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states on two benchmark data sets and compared their Schwarz weights. For the Andersson et al. (2017) data set, I expected the highest weights for the three-state models in the image condition. For the moving dots and video condition, I predicted the highest weights for the four-state models. Figure X shows the Schwarz weights for different subjects and models in the image condition. For all subjects, the five-state model displays the highest weight, suggesting that it is the most likely to have generated the data. In the moving dots condition, the five-state model has the highest weights for most subjects, but the one-, three, and four-state models exhibit the highest weights for two subjects each (see Figure X). The video condition has the same pattern as the image condition, as the five-state model consistently has the highest weights (see Figure X). In contrast to my hypothesis, applying different HMMs to the Andersson et al. data suggests that the five-state model has most likely generated the data. Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods.

Ehinger data set

@@

The hypothesis that HMMs improve the classification performance of eye movement events was investigated by comparing gazeHMM against other algorithms. Therefore, I applied gazeHMM with four events to the Andersson et al. (2017) data set. First, I calculated the RMSD [^2] as described in the original article by Andersson et al. for all algorithms included in the study plus gazeHMM. Table \@ref(tab:tab-rmsd-4-fix) shows that in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations due to a large amount of very short fixations. For saccades, gazeHMM has the fourth lowest RMSD of the algorithms in the image and video conditions but the second highest in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)). Here, gazeHMM overestimated the number and duration of saccades (either slightly or substantially). Moreover, gazeHMM underestimated the number and duration of PSOs in the image and moving dots conditions, leading to the highest and intermediate RMSD among the competitors, repsectively (see Table \@ref(tab:tab-rmsd-4-pso)). In the video condition the duration was slightly overestimated but the duration was underestimated with the highest RMSD. No other algorithm was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than among the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)). In sum, except for saccades and PSOs, gazeHMM performed worse with regard to the duration of classified events compared to the other algorithms.

@@

[^2]: In my proposal, I declared to compute the RMSE instead of RMSD. However, since I conducted an in sample comparison, the term deviation is more appropriate. Andersson et al. (2017) also referred to it as RMSD.

@@

(ref:plot.disag) Disagreement ratio between algorithms and human coders for different conditions. gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.

```{r plot-disag, fig.cap="(ref:plot.disag)"}

@@

Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms with regard to RMSD, sample-to-sample agreement, or disagreement ratio. It rather fared better or worse than most algorithms for different events and conditions. Especially for saccades, the agreement was relatively high and improved upon the IHMM algorithm included in the study.  
Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data might be more appropriate and yield better validation results. Table X shows the RMSD between gazeHMM and human coders only for the image condition. For fixations, gazeHMM shows a comparably low RMSD and only two other algorithms have lower values. Oppositely, saccades had the highest RMSD among the compared algorithms. The same result was found for PSOs.  
The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kapp-3) (only for the image condition). For fixations, the absolute kappa indicated substantial agreement and was high compared to the other algorithms. The absolute kappa for saccades corresponded to moderate agreement. Compared to the other algorithms, it was rather low. For PSOs, the absolute agreement was fair, but intermediate with regard to the other algorithms.  
Figure \@ref(fig:plot-disag) illustrates the disagreement ratio between gazeHMM with three events and the human coders compared to the other algorithms. In the image condition, the disagreement ratio was slightly lower for gazeHMM than for all the other algorithms.
Summing up, using gazeHMM with three states increased the agreement to human coding with regard to fixation durations and samples classified as fixations. However, for saccades and PSOs the agreement decreased.

>>>>

floatsintext      : no
figurelist        : yes
tablelist         : yes
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

@@

## Eye movement metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, the goal is to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007, and @Zemblys2018]. This can be problematic because these parameters are often set without theoretical justification and they differ substantially between metrics. Belonging to the most basic metrics, which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [similar to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur after high velocity saccades [@Larsson2013]. They are specifically distinguished by their change in direction clustered around 180° [@Pekkanen2017]. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015].

@@

(ref:filter-angle) Example data from @Andersson2017 displayed as sample-to-sample angle (in rad) over time (in s). Line colors indicate whether the positional data has been filtered before computing the sample-to-sample angle (red) or not (blue). For this example, I used a Butterworth filter of order three and a normalized cutoff frequency of 0.3. The full line segment marks where potential oscillations are filtered out. The dashed line segment illustrates where potentially random change in angle is filtered out.

@@

Instead of filtering before computing the metrics, I decided to combine both in one step. Previous algorithms have used two methods that both filter and compute derivatives of the gaze position (i.e., velocity and acceleration). @Houpt2018 compute the first and second discrete derivative (similar to a Sobel and Laplace filter, respectively) of the gaze position to be used by their algorithm. Similarly, @Holmqvist2010 use a Savitzky-Golay (SG) filter @Savitzky1964 to estimate velocity and acceleration from a gaze signal. Figure \@ref(fig:filter-comparison) displays velocity and acceleration signals obtained by both methods. They follow a similar trend, but the SG velocity signal is less noisy and the acceleration signal is consequently lower than the Laplace acceleration signal. Therefore, I chose to implement the SG filter to compute velocity and acceleration signals, because it filters out more noise than the discrete derivatives but still preserves the edges in the signal to distinguish between events. To preserve motor noise in the sample-to-sample angle signal, gazeHMM computes the absolute angle with the backward difference and the first discrete derivative as the forward difference (see Technical Appendix).

(ref:filter-comparison) Example data from @Andersson2017 displayed as velocity (deg/s) and acceleration (in deg/s$^2$) over time (in s). Line colors indicate which filter has been used to derive the signal. Savitzky-Golay filters were applied with order three and length five (corresponding to 10 ms).

@@

## The generative model
The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM). It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits. Thus, users can choose wether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth puruits.  
In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model [@Visser2019]. In gazeHMM, the response model consists of three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal. The response variables are treated as conditionally independent on the states. I acknowledge that conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated). This step was merely taken to keep the HMM simple and identifyable.  
Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them). However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation) and due to noise above zero. Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean. Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties. In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose. It has to be noted that the gamma was choosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von-Mises distributions (with a mean and concentration parameter) and a uniform distribution. Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$) which should lead to rather symmetric distributions. Moreover, the uniform distribution can distinguish fixations from the other events.  
The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, only parameter intercepts were estimated.

## Missing data and blinks
HMMs can estimate parameters and hidden states despite missing data. They either treat data as missing at random or as missing depending on states [@Visser2019]. In gazeHMM, I assume data to be missing at random. Most of the missing data in eye movement classification are due to blinks. The user can indicate in gazHMM which samples should be labeled as blinks (other missing samples are treated as noise). Often, eye-trackers record a few samples with unreasonably high velocity and acceleration before losing the pupil signal when a blink occurs. Since these samples could distort the classification of saccades in the HMM, I decided to remove them heuristically. Before classifying the samples, gazeHMM sets all samples within 50 ms before and after blink samples as missing. The window of 50 ms was rather motivated empirically than theoretically and can bet set to any value the user considers appropriate.  
When evaluating the likelihood of the HMM given parameters and the joint density of response distributions for missing data, the model integrates over all possible values. The resulting density for missing samples is therefore set to one [@Visser2019].

@@

apa_table(data.frame("Event" = rep(c("Fixation", "Saccade", "PSO", "Smooth pursuit"), each = 3),
                     "Resp variable" = rep(c("Velocity", "Acceleration", "Angle"), 4),
                     "Range" = c("0-50", "0-50", "uniform", "50-1000", "50-500", "~0", "20-100", "10-90", paste("~", expression(pi), sep = ""), "20-100", "0-30", "~0")), 
          caption = "(ref:sim-ranges)",
          note = "(ref:sim-ranges-note)")

@@

(ref:sim-parameters-trans) Intervals and default parameter values for the response models of the HMM in the simulation
(ref:sim-parameters-trans-note) The transition probability for staying in the same state is denoted by $a_{i=j}$ and the probability for switching to a different state by $a_{i\neq j}$. The number of states in the model is denoted by *k*.

```{r, sim-parameters-trans}

sim.pars.trans <- as.data.frame(matrix(c("Interval", "-", "[.01,.99]", "1-a(i=j)/(k-1)",
                                           "Default", "1/k", "0.9", "0.1/(k-1)"), nrow = 2, byrow = T))

apa_table(sim.pars.trans,
          col.names = c("", "Initial\nstate\nprob.", "Trans. prob.\nsame state", "Trans. prob.\nother state"),
          caption = "(ref:sim-parameters-trans)",
          note = "(ref:sim-parameters-trans-note)")

(ref:sim-parameters-resp) Intervals and default parameter values for the response models of the HMM in the simulation
(ref:sim-parameters-resp-note) Shape parameters are denoted by $\alpha$, scale parameters by $\beta$, mean parameters by $\mu$, and concentration parameters by $\kappa$. The default values for the uniform distribution in state one were min = 0 and max = $2\pi$.

```{r, sim-parameters-resp}

sim.pars.resp <- as.data.frame(t(matrix(c("[1,5]", "[0.1,0.6]", "[1,5]", "[0.05,0.25]", "-", "-", 
                                  "3", "0.35", "3", "0.25", "-", "-",
                                  "[1,5]", "[5,15]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
                                  "3", "10", "3", "3", "0", "1",
                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[1,5]", "-", "1/[0.1,10]",
                                  "3", "1", "3", "3", paste(expression(pi), sep = ""), "1",
                                  "[1,5]", "[0.5,1.5]", "[1,5]", "[0.05,0.25]", "-", "1/[0.1,10]",
                                  "3", "1", "3", "0.15", "0", "1"), nrow = 6, byrow = F)))

apa_table(cbind(rep(c("Interval", "Default"), 4), sim.pars.resp),
          col.names = c("", "Shape", "Scale", "Shape", "Scale", "Mean", "Concentration"),
          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
          stub_indents = list("State 1" = c(1, 2), "State 2" = c(3, 4), "State 3" = c(5, 6), "State 4" = c(7, 8)),
          caption = "(ref:sim-parameters-resp)",
          note = "(ref:sim-parameters-resp-note)")

@@

```{r load sim results and helper functions, include=FALSE}

@@

## Validation
### Starting values and model fitting
I applied gazeHMM on two benchmark data sets to test my hypotheses. For both data sets, I used the same starting values (see Table X) to estimate the parameters of the HMM. To improve the fitting of the gamma distributions, velocity and acceleration signals were scale down by factor 100 (and so were the starting values for their gamma distributions). The algorithm was applied separately for every participant and every condition. For the data set by Andersson et al. [-@Andersson2017], the algorithm was successfully fitted for every participant and condition.

(ref:tab-starting-values) Starting values for fitting gazeHMM to benchmark data sets
(ref:tab-starting-values-note) Starting values for velocity and acceleration signals are shown before scaling down by factor 100. Values for event 5 were chosen so that they approximately match plausible distributions for microsaccades.

```{r tab-starting-values}

tab.start <- as.data.frame(matrix(c(10, 10, 10, 10, 0, 2*pi,
                      50, 50, 50, 50, 0, 10,
                      50, 50, 50, 50, pi, 10,
                      20, 20, 20, 20, 0, 10,
                      20, 20, 50, 50, 0, 10), nrow = 5, byrow = T))

colnames(tab.start) <- c("Shape", "Scale", "Shape", "Scale", "Mean", "Concentration")
rownames(tab.start) <- c("Fixation", "Saccade", "PSO", "Pursuit", "Event 5")

apa_table(as.data.frame(tab.start),
          col_spanners = list(Velocity = c(2, 3), Acceleration = c(4, 5), "Rel. angle" = c(6, 7)),
          caption = "(ref:tab-starting-values)",
          note = "(ref:tab-starting-values-note)")

### Model comparison

@@

To examine whether HMMs are accurately describing eye movements, I applied them with one, two, three, four, and five states on two benchmark data sets and compared their Schwarz weights. For the Andersson et al. [-@Andersson2017] data set, I expected the highest weights for the three-state models in the image condition. For the moving dots and video condition, I predicted the highest weights for the four-state models. Figure X shows the Schwarz weights for different subjects and models in the image condition. For all subjects, the five-state model displays the highest weight, suggesting that it is the most likely to have generated the data. In the moving dots condition, the five-state model has the highest weights for most subjects, but the one-, three, and four-state models exhibit the highest weights for two subjects each (see Figure X). The video condition has the same pattern as the image condition, as the five-state model consistently has the highest weights (see Figure X). In contrast to my hypothesis, applying different HMMs to the Andersson et al. data suggests that the five-state model has most likely generated the data. Except for the moving dots condition, there was little variation in the Schwarz weights, indicating large differences in the model likelihoods. The model comparison for the Ehinger et al. [-@Ehinger2019] data set yielded similar results (see Appendix).

@@

The hypothesis that HMMs improve the classification performance of eye movement events was investigated by comparing gazeHMM against other algorithms. Therefore, I applied gazeHMM with four events to the Andersson et al. (2017) data set. First, I calculated the RMSD [^1] as described in the original article by Andersson et al. for all algorithms included in the study plus gazeHMM. Table \@ref(tab:tab-rmsd-4-fix) shows that in all three conditions, gazeHMM has a higher RMSD than all the the other algorithms for classified fixations due to a large amount of very short fixations. For saccades, gazeHMM has the fourth lowest RMSD of the algorithms in the image and video conditions but the second highest in the moving dots condition (see Table \@ref(tab:tab-rmsd-4-sac)). Here, gazeHMM overestimated the number and duration of saccades (either slightly or substantially). Moreover, gazeHMM underestimated the number and duration of PSOs in the image and moving dots conditions, leading to the highest and intermediate RMSD among the competitors, repsectively (see Table \@ref(tab:tab-rmsd-4-pso)). In the video condition the duration was slightly overestimated but the duration was underestimated with the highest RMSD. No other algorithm was designed to classify smooth pursuits, but gazeHMM classified a large number of short smooth pursuits, causing a substantially higher RMSD than among the human coders (see Figure \@ref(tab:tab-rmsd-4-sp)). In sum, except for saccades and PSOs, gazeHMM performed worse with regard to the duration of classified events compared to the other algorithms.

@@

[^1]: In my proposal, I declared to compute the RMSE instead of RMSD. However, since I conducted an in sample comparison, the term deviation is more appropriate. Andersson et al. (2017) also referred to it as RMSD.

@@

(ref:plot-disag) Disagreement ratio between algorithms and human coders for different conditions. gazeHMM-3 classified three and gazeHMM-4 classified four events. Figure adapted from @Andersson2017.

```{r plot-disag, fig.cap="(ref:plot-disag)"}

@@

Contrary to my second hypothesis, applying gazeHMM to a benchmark data set revealed that it did not outperform all other algorithms with regard to RMSD, sample-to-sample agreement, or disagreement ratio. It rather fared better or worse than most algorithms for different events and conditions. Especially for saccades, the agreement was relatively high and improved upon the IHMM algorithm included in the study.  
Even though it was not confirmed by the model comparison, applying gazeHMM with three states to static data might be more appropriate and yield better validation results. Table \@ref(tab:tab-rmsd-3) shows the RMSD between gazeHMM and human coders only for the image condition. For fixations, gazeHMM shows a comparably low RMSD and only two other algorithms have lower values. Oppositely, saccades had the highest RMSD among the compared algorithms. The same result was found for PSOs.  
The agreement between gazeHMM with three states and the human coders measured by Cohen's kappa is displayed in Table \@ref(tab:tab-kappa-3) (only for the image condition). For fixations, the absolute kappa indicated substantial agreement and was high compared to the other algorithms. The absolute kappa for saccades corresponded to moderate agreement. Compared to the other algorithms, it was rather low. For PSOs, the absolute agreement was fair, but intermediate with regard to the other algorithms.  
Figure \@ref(fig:plot-disag) illustrates the disagreement ratio between gazeHMM with three events and the human coders compared to the other algorithms. In the image condition, the disagreement ratio was slightly lower for gazeHMM than for all the other algorithms.
Summing up, using gazeHMM with three states increased the agreement to human coding with regard to fixation durations and samples classified as fixations. However, for saccades and PSOs the agreement decreased.

(ref:tab-rmsd-3) Mean and standard deviation of event durations, number of events, and RMSD between event durations classified by algorithms and human coders for the image condition
(ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.

```{r tab-rmsd-3}

tab.rmsd.3 <- lapply(A2017.rmsd[[2]][[2]], function(x) {

  df <- as.data.frame(t(as.matrix(x)))

  rownames(df)[3] <- "gazeHMM"

  out <- cbind(algorithm = rownames(df), df)

  return(out)
}) %>% reduce(left_join, by = "algorithm")

apa_table(tab.rmsd.3,
          col.names = c("Algorithm", rep(c("Mean", "SD", "Events", "RMSD"), 3)),
          col_spanners = list("Fixations" = c(2, 5), "Saccades" = c(6, 9), "PSOs" = c(10, 13)),
          font_size = "tiny",
          caption = "(ref:tab-rmsd-3)",
          note = "(ref:tab-rmsd-3-note)")
